# Larval response to parental low pH exposure in Pacific oysters (*Crassostrea gigas*)

## Abstract

As negative effects of ocean acidification are experienced by coastal ecosystems, there is a growing trend to investigate the effect ocean acidification has on multiple generations. Parental exposure to ocean acidification has been shown to induce larval carryover effects, but whether or not acute exposure to a stressor as an adult can influence the larval generation long after the stress has been removed has yet to be tested. To assess how a temporary exposure to experimental ocean acidification affects the ecologically and commercially relevant Pacific oyster (*Crassostrea gigas*), adult oysters were exposed to either low pH (7.31 ± 0.02) or ambient pH (7.82 ± 0.02) conditions for seven weeks. Oysters were then held for eight weeks in ambient conditions, and subsequently reproductively conditioned for four weeks at ambient pH. After conditioning, oysters were strip-spawned to create four families based on maternal and paternal ocean acidification exposure. The number of D-hinge larvae were counted eighteen hours post-fertilization. A sex-specific broodstock response was observed, where female exposure to low pH conditions resulted in fewer D-hinge larvae. This study demonstrates that the effects of ocean acidification can last beyond the time from when the environmental perturbation is experienced. Broadening the understanding of environmental memory will be valuable when considering organismal ability to persist in the face of environmental change.

## Introduction

Determining how parental exposure to ocean acidification carries over into early larval stages is important for understanding cumulative effects of climate-related environmental change. Gametogenesis is a key period during which parental exposure to ocean acidification can influence offspring [@Donelson2018]. Several studies exposing Sydney rock oysters (*Saccostrea glomerata*) to high pCO~2~ conditions (856 µatm, pHNBS 7.89-7.90) during reproductive conditioning identified positive larval carryover effects [@Parker2012; @Parker2015; @Parker2017]. Specifically, larvae from parents exposed to low pH conditions were larger and developed faster in acidified conditions compared to those from parents reared in ambient pH conditions [@Parker2012; @Parker2015; @Parker2017]. Conversely, similar experiments conducted with northern quahog (= hard clam; *Mercenaria mercenaria*) and bay scallops (*Argopecten irradians*) demonstrated negative larval carryover effects [@Griffith2017]. Larvae from adult *A. irradians* and *M. mercenaria* exposed to low pH were more sensitive to acidified conditions than those spawned from parents exposed to ambient pH during reproductive conditioning [@Griffith2017]. These studies demonstrate the importance of parental exposure during reproductive conditioning (late-stage gametogenesis) on offspring. 

As the Pacific oyster (*Crassostrea gigas*; Thunberg, 1793) is a commercially and ecologically relevant species in much of the world, several research efforts have identified consequences of ocean acidification for distinct *C. gigas* life stages. While fertilization still occurs under near-future ocean acidification conditions [@Kurihara2007; @Havenhand2009; @Boulais2018], fertilization success in acidified conditions is variable between *C. gigas* populations [@Parker2010; @Barros2013]. Researchers have found that larvae experience developmental delays and reduced shell growth when exposed to experimental ocean acidification conditions [@Kurihara2007; @Gazeau2011; @Timmins-Schiffman2013; @Waldbusser2014]. Natural upwelling-induced ocean acidification conditions also reduced larval production and growth in a hatchery setting [@Barton2012]. Ocean acidification hampers protein expression in larvae, especially for proteins related to calcification and cytoskeleton production [@Dineshram2012]. During metamorphosis, oyster larvae experience down-regulation of proteins related to energy production, metabolism, and protein synthesis [@Dineshram2016]. Adult *C. gigas* calcification rates decrease as seawater pCO~2~ increases [@Gazeau2007], with oysters grown at 2800 µatm displaying significantly lower fracture toughness than oyster shells from ambient conditions [@Timmins-Schiffman2014]. Exposure to ocean acidification also affects adult antioxidant responses, carbohydrate metabolism, transcription, and translation protein pathways [@Timmins-Schiffman2014]. Predator-prey interactions can also change under experimental ocean acidification [@Wright2018]. There is limited evidence, however, of how ocean acidification influences Pacific oysters across multiple generations.

The current study is the first to discern how exposure to experimental ocean acidification prior to reproductive conditioning affects larval abundance in *C. gigas*. This experiment not only describes how isolated exposure to low pH during early gametogenesis influences larvae, but also provides information on the effects of acute pH exposure on adult gonad morphology. Additionally, the study demonstrates how environmental perturbation experienced before reproductive maturity affects the subsequent generation, even if the stressor is long-removed.

## Methods

### Experimental overview

Experimental trials were conducted at the Kenneth K. Chew Center for Shellfish Research and Restoration at the National Oceanic and Atmospheric Administration (NOAA) Manchester Field Station (47°34'09.1"N 122°33'19.0"W, Manchester, Washington, USA) in 2017. Adult hatchery-raised *C. gigas* (average shell length = 117.46 ± 19.16 cm) were acclimated in the facility for 10 days, then exposed to either low or ambient pH conditions for 48 days (Figure 1). After pH exposure, oysters were held at ambient pH and water temperature conditions for 90 days. Oysters underwent reproductive conditioning for 22 days, then strip-spawned. D-hinge larvae were counted eighteen hours after fertilization occurred.

### Experimental pH exposure

The experimental system consisted of a 1,610 liter storage tank that fed two 757 liter header tanks. Water from Clam Bay, WA was pumped through a sand filter, then UV-treated. The UV-treated water passed through a set of three sock filters (100 µm, 50 µm, and 25 µm) and a degassing column. Once degassed, water passed through three more sock filters (25 µm, then 10 µm, and 5 µm) before entering the storage tank. The storage tank was outfitted with an off-gas vent and pump to recirculate water such that CO~2~ in the water could be equilibrated with atmospheric CO~2~. Equilibrated water flowed into the two header tanks, each of which fed three 50L flow-through (1.2 L/min) experimental tanks (six experimental tanks total). For all header and experimental tanks, pH in header and experimental tanks was continuously monitored using Durafet pH probes (Honeywell Model 51453503-505) and an AVTECH system. The addition of CO~2~ in the low pH header tank was controlled using a solenoid valve. A Dual Input Analytical Analyzer (Honeywell Model 50003691-501) automatically mediated solenoid injections. A CO~2~ airline with a back pressure of 15 psi, controlled with a regulator, injected CO~2~ into the low pH header tank every 180 seconds with an injection duration of 0.4 seconds. Injections only occurred if real-time pH from the Durafet was above pH 7.22. A venturi injector connected to the ambient water line mixed ambient pH water with CO~2~-rich water to lower pH. There were no CO~2~ injections in the ambient header tank.

Prior to the pH exposure trial, twenty randomly selected *C. gigas* were lethally sampled to assess gonadal status (see *Histological analysis*). Oysters were placed in each flow-through experimental tank in ambient water conditions and exposed to ambient or low pH conditions for seven weeks. Each treatment consisted of 3 tanks, each with 20 oysters. All experimental tanks received algae from a common reservoir. The algal tank contained 300-500 mL of Shellfish Diet 1800® (Reed Mariculture) diluted in 200L of ambient pH seawater [@Helm2004]. Algae was continuously dosed to oyster experimental tanks using an Iwaki Metering Pump. Algal lines were cleaned twice weekly, and experimental tanks were fully drained and cleaned once a week.

### Seawater chemistry analysis

Twice a week, water samples (1L) were collected from each header and oyster experimental tank. For each sample, salinity was measured with a Bench/Portable Conductivity Meter (Model 23226-505, VWR), pH (mV) was measured with a Combination pH Electrode (Model 11278-220, Mettler Toledo), and temperature (ºC) was measured using a Traceable Digital Thermometer (Model 15-077, Fisher). To calibrate the pH probe, a Tris buffer (0.08 M, 28.0 salinity) was prepared using 0.3603 mol of NaCl (J.T. Baker), 0.0106 mol of KCl (Fisher Scientific), 0.0293 mol MgSO4-(H2O)7 (Fisher Scientific), 0.0107 mol of CaCl2-2(H2O) (MP Biomedicals), 0.0401 HCl (J.T. Baker), and 0.0799 mol of Tris base (Fisher Scientific). Deionized water was added for a final volume of 1L. Salinity, temperature, and pH measurements for the Tris buffer were obtained at five temperatures before measuring samples to generate a standard curve. This standard curve was used to calibrate the pH electrode and convert measured millivolts to pH units.

For total alkalinity measurements, duplicate seawater samples (250 mL) were collected from experimental tanks twice weekly and dosed with mercuric chloride (50 µL of 0.18 M solution) to preserve samples [@Bandstra2006]. Samples from days 5, 33, and 48 were run on a T5 Excellence titrator (Mettler Toledo) to determine alkalinity. Salinity from discrete samples was used to calculate total alkalinity, using the `seacarb` library in R [@Gattuso2018]. Calculated pH, total alkalinity, temperature, and salinity were also used in `seacarb` to calculate in situ pH, pCO~2~, dissolved organic carbon (DIC), calcite saturation (Ωcalcite), and aragonite saturation (Ωaragonite) for days 5, 33, and 48. R code used to calculate water chemistry parameters is available [@Venkataraman2018].

### Histological analysis

Twenty randomly selected *C. gigas* were lethally sampled before pH exposure for histological analyses. On the last day of low pH exposure, ten oysters from each treatment — randomly selected from each tank — were also lethally sampled to assess gonadal status. For each sampled oyster, a piece of gonad tissue was cut and placed in a histology cassette. Gonad tissue in cassettes was fixed for histology using PAXgene Tissue FIX and STABILIZER and sent to Diagnostic Pathology Medical Group, Inc. (Sacramento, CA) for staining with hematoxylin and eosin and slide preparation. Tissues exposed to ambient pH were confounded during processing, preventing any tank identification. Maturation state and organism sex was evaluated histologically at 40x magnification [@Fabioux2005; @Enríquez-Díaz2008].

### Reproductive conditioning

Following seven weeks of low pH exposure, oysters were returned to a common garden and maintained at ambient pH conditions for eight weeks. Afterwards, oysters were reproductively conditioned. Water temperatures and food quantity are known to regulate the timing, speed, and intensity of gametogenesis in *C. gigas* [@Enríquez-Díaz2008]. Conditioning protocol was modeled after standard hatchery practices (Molly Jackson, Broodstock Manager at Taylor Shellfish, pers. comm., June 2017). Water temperature was raised from ambient conditions (13ºC) to 23ºC over three weeks (1ºC/2 days), since optimal temperature for *C. gigas* gametogenesis is between 18ºC and 26ºC [@Parker2010]. Conditions were maintained at 23ºC for one more week prior to spawning. During conditioning, *C. gigas* were fed 700-800 mL of Shellfish Diet 1800® daily [@Helm2004]. 

### Strip spawning and larval rearing

After reproductive conditioning, all surviving oysters were prepared for strip spawning. A sample of gonad from each individual was assessed for presence of active sperm or eggs using a microscope at 10x magnification. Only *C. gigas* with active sperm or eggs were used for crosses (n~male, low~ = 6, n~female, low~ = 22, n~male, ambient~ = 6, n~female, ambient~ = 26). Presence of mature gametes and ripe oysters indicated that oysters were in good condition and not affected by use of Shellfish Diet 1800® instead of live algae during reproductive conditioning. For each treatment (low pH and ambient conditions), one gram of mature gonad from each ripe female was pooled. The number of eggs in both the ambient and low pH pools were counted to determine the number of eggs used for parental crosses. Parental crosses were created using 210,000 eggs from the female egg pools and sperm (200 µL) from individual males. 

Four half-sibling families were created based on parental pH exposure: low pH female (pool) x low pH male, low pH female (pool) x ambient pH male, ambient pH female (pool) x low pH male, and ambient pH (pool) female x ambient pH male. These crosses were conducted using pooled eggs from either low pH or ambient pH females, and sperm from one of six males within each pH treatment (e.g. low pH female pool x low pH male-01, low pH female pool x low pH male-02, ... low pH female pool x low pH male-06), totaling 24 crosses. All crosses were performed in duplicate, resulting in 48 separate fertilization events. 

Fertilization was carried out in plastic beakers (1L) for 20 minutes with static 23ºC filtered seawater (1 µm) in ambient pH conditions. After confirming polar body formation, beaker contents were transferred to larger plastic tanks (19L) with aerated, static 23ºC filtered seawater (1 µm) for eighteen hours of incubation. Duplicate containers were combined eighteen hours post-fertilization, and D-hinge larvae were counted for each cross (n= 24).

### Statistical analyses

Differences in in situ pH, total alkalinity, pCO~2~, DIC, Ωcalcite, and Ωaragonite between pH treatments were evaluated with a one-way ANOVA. Because tissue samples were confounded during histological processing, a binomial GLM model was used to compare gonad maturation between pH treatments. Differences in sex ratios between pH treatments were evaluated using a chi-squared test of homogeneity. To identify differences in D-hinge larval counts, a linear mixed model was used, with sire and female egg pool as random effects. Differences in D-hinge larval counts by female treatment were assessed using a similar linear mixed model, with only sire as a random effect. Normality of data, as well as independence and homoscedasticity, were verified visually. All statistical analyses were carried out in R (Version 3.4.0). R Scripts are available in the supplementary Github repository [@Venkataraman2018].



##Results

$\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ values did not vary by sex (p > 0.05, Figure \@ref(fig:dist) or size for the individuals sampled (p > 0 .05; Figure \@ref(fig:Length)). Spatial variation in harbor seal $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values were observed on subregional scales. $\delta^{15}N_{Phe}$ values were similar for harbor seals in the northern Gulf of Alaska (11.9  ± 2.9, mean ± 1SD), southeast Gulf of Alaska (10.8 ± 1.7), and coastal Washington (11.3 ± 1.9). The eastern Bering Sea had significantly higher $\delta^{15}N_{Phe}$ values compared to other subregions (15.2 ± 1.8) followed by the Salish Sea (12.2 ± 2.3) which had similar $\delta^{15}N_{Phe}$ values compared to the northern Gulf of Alaska (Figure \@ref(fig:dist), Table \@ref(tab:ttestN)). $\delta^{13}C$ values varied by subregion (p < 0.05) with the exception of the Gulf of Alaska, where the northern (-14.6 ± 0.9) and southeast (-14.4 ± 1.1) subregions were not significantly different, and the eastern Bering Sea (-13.4 ± 0.9) and coastal Washington (-13.6 ± 0.9) were not significantly different (Figure \@ref(fig:dist), Table \@ref(tab:ttestC)). The variation between subregions appeared to follow a latitudinal gradient, where harbor seal mean $\delta^{13}C$ values were most enriched in $^{13}C$ in the Salish Sea (-12.2 ± 1.5), became more depleted from coastal Washington and into the Gulf of Alaska (Table \@ref(tab:ranges)).
	
The relationship between harbor seal $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values also varied on subregional scales. There was positive linear association between harbor seal $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values in the combined northeast Pacific basin and Bering Sea model with a slope of 0.12 (Figure \@ref(fig:hiermod)A). For the hierarchical subregion model, the eastern Bering Sea and coastal Washington demonstrated similar relationship, with slopes of 0.08 (95% CI [0.05, 0.11]) and 0.07 (95% CI [0.05, 0.09]) respectively. Similarly, harbor seals in both Gulf of Alaska subregions demonstrated comparable coupling of $\delta^{15}N_{Phe}$ and $\delta^{13}C$, with slopes of 0.13 (95% CI [0.11, 0.14]) for the northern subregion and 0.12 (95% CI [0.10, 0.14]) for the southeastern subregion. Salish Sea harbor seals had a distinct relationship between $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values relative to other subregions with a slope of only 0.02 that was not significantly different from 0 (95% CI [0.0, 0.04]) (Figure \@ref(fig:hiermod)B).
	
For both $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values there was substantial support for models including environmental indices rather than null or subregion only models. The relationship between environmental indices and harbor seal $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values in the northeast Pacific varied on regional scales. For Washington, the best model to predict harbor seal $\delta^{15}N_{Phe}$ values included Columbia River discharge in high flow months, summer upwelling, and subregion. There was substantial model uncertainty for $\delta^{15}N_{Phe}$ values in the Washington region, however 90% of model weight supported the inclusion of Columbia River discharge (Figure \@ref(fig:coefres)A). The model for harbor seal $\delta^{13}C$ values with the most support indicated a positive association between PDO, spring upwelling, and freshwater discharge in the Washington region (Figure \@ref(fig:coefres)B). In the Gulf of Alaska, the summer upwelling model had the most support as a predictor of harbor seal $\delta^{15}N_{Phe}$ values with some model support for inclusion of the NPGO (North Pacific Gyre Oscillation), although the coefficients for this covariate did not differ substantially from 0 (Figure \@ref(fig:coefres)C). The best model for harbor seal $\delta^{13}C$ values for the Gulf of Alaska included subregion, PDO (Pacific Decadal Oscillation), and NPGO (Figure \@ref(fig:coefres)D). In contrast to Washington, the Gulf of Alaska models supported a negative association between δ13C values and PDO. The null model for $\delta^{15}N_{Phe}$ values in the eastern Bering Sea had the most support (Figure \@ref(fig:coefres)E). Lack of model support for environmental covariates in the eastern Bering Sea may have been a result of the small sample size in the region. Cross-shelf wind was included as a predictor in the best model (Figure \@ref(fig:coefres)F) for $\delta^{13}C$ values in the eastern Bering Sea and was supported by 76% of the model weight. 
	
PDO and Kuskokwim river discharge during high flow months were found to be highly collinear (VIF > 10) and PDO was omitted from the candidate model set for the eastern Bering Sea analysis. All other models containing multiple environmental predictors with relative support had variance inflation factors of less than 2 indicating only moderate collinearity across covariates. Model residuals for the best models did not show trends through time (Figure \@ref(fig:linresid)). This indicates that there were no trends associated with other potential ecosystem changes, such as harbor seal foraging strategy for example, after accounting for ocean condition. Model results did not change when using $\delta^{13}C$ data that were not corrected for the regional Seuss Effect.
	
The GPDFA analysis showed temporal synchronies and shared trends across environmental conditions and stable isotope values in the northeast Pacific. In the Gulf of Alaska, the data supported three latent trends (Figure \@ref(fig:GDFAAK)). Both $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values had the highest loadings for trend 1, which showed an increase starting in 1965 until 1980 followed by the trend oscillating at approximately 25% above the long-term average. The harbor seal $\delta^{15}N_{Phe}$ values for the southeast subregion, harbor seal $\delta^{13}C$ values, and spring upwelling had negative loadings on trend 1; loadings of $\delta^{15}N_{Phe}$ values were generally weaker relative to loadings of $\delta^{13}C$ values. For the other two trends (2-3), loadings were clustered by environmental driver category. Latent trend 2 oscillated around the long-term average and was uninformative. Trend 3 was below average starting in 1985 with strong loadings for climate time series, spring and summer upwelling, and discharge in high flow months (Figure \@ref(fig:GDFAAK)). Annual discharge, autumn upwelling, Oceanic Niño Index and Northern Oscillation Index did not demonstrate strong loadings for any trend. In Washington, there was support for two latent trends. Latent trend 1 shows a rapid increase in the 1940’s to 25% above the long-term mean then a gradual decline until 1986 to approximately 40% below the long-term mean, with values below the mean starting in 1977 (Figure \@ref(fig:GDFAWA)). Trend loadings for harbor seal $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values were stronger for coastal seals and trend 1 had stronger loadings for freshwater discharge than trend 2. Trend 2 had strong loadings for $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values for both Salish Sea and coastal Washington harbor seals. Trend 2 oscillated above and below the long-term mean and had large loadings for sea surface temperature, summer upwelling, Fraser River discharge and climate indices (Figure \@ref(fig:GDFAWA)).

To validate this approach of applying a 1-year lag, 0- and 2-year lags were also applied to the best models of each region an compared to the 1-year lag using AIC. Additionally, month was tested as a smoothed predictor with 12 knots for stable isotope data in Washington samples using a generalized additive model (GAM). Support for a significant smoothing term would identify and seasonality in the data, which would be expected if tissue turnover time is less than a year. A one-year tissue turnover time was confirmed as a suitable assumption for harbor seal bone collagen, as 0- and 2- year lags had similar or less model support. There was no support for a smoothing effect by month in generalized additive models of $\delta^{15}N_{Phe}$ and $\delta^{13}C$ which would have indicated any seasonal variability in isotope composition and thus a turnover time of less than or greater than a year (p < 0.05; Figure \@ref(fig:Length)) and thus a 1-year lag was applied to isotope data for all temporal analyses.

##Discussion

We analyzed bone collagen $\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ values from harbor seal museum specimens collected between 1928 and 2014 as indices of change in food web assimilated nitrogen and carbon. Based on previous research [@Sherwood2014; @delaVega2019; @Lorrain2020], we interpret $\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ values as primarily representing nitrogen and carbon resource utilization, and growth and community composition of primary producers at the base of the food web. Our data show the relationship between indices of primary production and nitrogen resources assimilated into food webs varies regionally across the northeast Pacific. By pairing these data with environmental time series data, we provide new insights into large scale environmental forcing that impacts the base of the food web and is transferred to higher trophic levels. Specifically, oceanic conditions associated with climate regimes and upwelling explain significant temporal variation in $\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ values of coastal predators in northeast Pacific (Figure \@ref(fig:coefres); Figure \@ref(fig:linresid)). This analysis demonstrates $\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ values are useful indicators of resources assimilated by coastal food webs.  

###Spatial variation in stable isotope indices

The geographically widespread association between harbor seal $\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ values indicates food web assimilated primary production is coupled with nitrogen resources in most regions of the northeast Pacific, with the Salish Sea as a notable exception (Figure \@ref(fig:hiermod)). Short-term studies in coastal Washington showed phytoplankton respond considerably to nitrogen inputs and are frequently nitrogen limited [@Dortch1989; @Kudela2009]. Similarly, short term studies of the inner Gulf of Alaska shelf demonstrated primary production is generally nitrogen limited, and size, growth rates, and community composition are all tightly coupled with nutrient availability [@Strom2006]. A significant relationship between bulk $\delta^{15}N$ and $\delta^{13}C$ values was also observed in the tissues of some gorgonian corals over the same time period in coastal Gulf of Alaska [@Williams2007]. Given the evidence of nitrogen limitations and its relationship with phytoplankton growth and community composition in these coastal environments, the association between $\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ values could be the result of nitrogen limiting growth at the base of the food web. Alternatively, the $\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ coupling could be driven by covariance with an untested environmental variable that impacts most of the northeast Pacific but not the Salish Sea.
	 
The coastal Washington and the Salish Sea food webs assimilate different nitrogen and carbon sources (Figure \@ref(fig:coefres) A & B). Salish Sea harbor seals have higher $\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ values compared to individuals on the outer coast, which is likely due to significant contributions of intertidal producers and the legacy of anthropogenic N in the Salish Sea food web. Intertidal macrophytes (seagrass and algae) have similar δ13C values (~ -10‰) compared to harbor seals in the Salish Sea, while other potential sources are much lower (i.e., marine derived sources ~ -20‰, terrestrial derived sources ~ -30‰) [@Conway2015; @Howe2015]. Incorporation of intertidal producers into the Salish Sea food web explains the difference in carbon stable isotope signatures between Salish Sea and coastal Washington harbor seals (~1.4‰, Figure \@ref(fig:coefres)A). However, it does not explain the higher $\delta^{15}N_{Phe}$ values (Figure \@ref(fig:coefres)B, Table \@ref(tab:ranges)). Surface nitrate was observed to be 8‰ – 12‰ off the coast of Washington in spring 1993 [@Wu1997] which was exceeded by harbor seals in both coastal Washington and the Salish Sea (Table \@ref(tab:ranges)). It is likely anthropogenically derived nitrogen sources contribute to the higher observed $\delta^{15}N_{Phe}$ values both directly and indirectly, particularly in the Salish Sea where harbor seal $\delta^{15}N_{Phe}$ values were up to 2.4‰ higher than coastal Washington seals. Wastewater treatment facilities and agriculture runoff contribute substantial amounts (~32%) of nitrogen in the Salish Sea (Mohamedali et al. 2011) and are enriched in $^{15}N$. In recent decades, Salish Sea waters have also been characterized by low dissolved oxygen and hypoxic events (PSEMP 2019) from human derived nitrogen loading. Anoxic conditions are conducive to denitrification, another potential indirect source of $^{15}N$ from human activities in the region. 

###Ocean condition and stable isotope indices

Washington state food webs exhibit environmentally induced changes in assimilated primary production and nitrogen sources. The isotope-ocean condition relationship in the region can be explained by introduction of terrestrial derived nutrients and climatically induced changes in phytoplankton community structure observed in previous studies [@Du2015; @Du2014; @Kudela2008]. For example, the PDO has been associated with phytoplankton community shifts between dinoflagellates and diatoms in the northern California Current [@Du2015]. Similarly, the phytoplankton community composition is distinct in the early (spring) upwelling season compared to the late (summer) upwelling season [@Du2014]. This could explain the inversely related associations between bulk $\delta^{13}C$ values and summer and spring upwelling (Figure \@ref(fig:coefres)B). Shifts in phytoplankton community structure are therefore a mechanism to explain the relationship between harbor seal bulk $\delta^{13}C$ values and ocean condition. In addition, freshwater discharge explains 16% of variation observed in both $\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ values in Washington. The Columbia River Plume introduces terrestrial derived nutrients, including nitrogen, and has been associated with increased primary production and fish production [@Kudela2008]. The covariation between $\delta^{15}N_{Phe}$, bulk $\delta^{13}C$, and discharge indicates isotopically distinct nitrogen resources introduced by freshwater discharge alters primary production which is then assimilated into the Washington food web, and ultimately harbor seals. 

In the eastern Bering Sea, our results suggest ice-born algae and $^{15}N$ enriched nitrogen from the inner shelf are important for supporting the coastal food web. Recent evidence supports that consumer $\delta^{15}N_{Phe}$ values reflect nitrate $\delta^{15}N$ values in the arctic (de la Vega et al. 2020). However, our $\delta^{15}N_{Phe}$ values from harbor seals of the eastern Bering Sea were high relative to previous studies of summer nitrate (5 to 9‰) [@Lehmann2005] and plankton nitrogen isotope signatures (6-12‰) [@Smith2002] from the outer and mid Bering Sea shelf. @Morales2014 subsequently found the stable isotope composition of nitrogen in diatoms ranged from 5-21‰ in late winter and early spring. These values also increased in association to sea ice with a positive shoreward gradient [@Morales2014]. The range of sea-ice algae $\delta^{15}N$ values observed by @Morales2014 are consistent with our observed $\delta^{15}N_{Phe}$ values in harbors seals (\@ref(tab:ranges)). Furthermore, the harbor seals in this study were located near the inner shelf in an area that has been partially covered by sea ice from January to May during the past century [@Stabeno2007]. Together this indicates ice algae as a significant contributor to the coastal food web. The disconnect between the $\delta^{15}N$ values of offshore nitrate [@Lehmann2005] and harbor seals also highlights the problem in assuming spatially and temporally discrete nitrate or phytoplankton measurements are representative of resources utilized by, and assimilated into, coastal food webs. Consumer $\delta^{15}N_{Phe}$ measurements by their nature represent the N assimilated into the food web and integrated over relatively long time scales, while discrete measurements of nitrate may be spatially or temporally biased. 

A short term (1998-2011) study of abiotic drivers in the Gulf of Alaska found chlorophyll-a anomalies were positive when downwelling favorable winds were low and had a negative relationship with sea level [@Waite2013]. Similarly, @Espinasse2020 found chlorophyll-a, SST, and sea level anomalies were the best predictors of carbon and nitrogen isotope data for secondary consumers over the past two decades. Our results agree with these studies as NPGO (an index of sea level) is negatively associated with both harbor seal $\delta^{15}N_{Phe}$ (Figure \@ref(fig:coefres)C) and $\delta^{13}C$ values (Figure \@ref(fig:coefres) D) in the Gulf of Alaska. Similarly, summer upwelling is positively associated with our $\delta^{15}N_{Phe}$ values (Figure \@ref(fig:coefres)C). Based on our results, these environmentally induced changes represent long-term ecosystem dynamics that extend beyond merely the base of the food web and ultimately impact resources assimilated by top predators. In addition, regional climate indices characterize nutrient and primary production assimilated annually into the food web better than sea surface temperature data alone. It is possible that other untested abiotic factors such as cross-shelf exchanges via eddy propagation or local wind stress [@Waite2013] may be important to food web assimilated nitrogen and primary production in the Gulf of Alaska. Regardless, local variability in upwelling and basin scale indices of sea surface height and temperature (i.e., NPGO) ultimately determine resource assimilation in Gulf of Alaska food web in which harbor seals forage.

By comparing consumer stable isotope values against environmental covariates across multiple sub basins we show environmental forcing on coastal food webs is regionally distinct. For example, climate indices (i.e., PDO) in the Gulf of Alaska were inversely associated with food web-assimilated primary production (Figure \@ref(fig:coefres)D, Figure \@ref(fig:GDFAAK) Trend 1-2) and positively associated in Washington (Figure \@ref(fig:coefres)B, Figure \@ref(fig:GDFAWA) Trends 1-2). This agrees with previous studies where the Pacific Decadal Oscillation has been associated with alternating salmon production in the northeast Pacific [@Mantua1997]. In cool phase years (i.e., 1947-1977), Washington stocks experience above average production and Alaska stocks experience below average production. Our results show that $\delta^{13}C$ values for Washington and Gulf of Alaska also indicate alternating primary production between the two regions in association with PDO. Surprisingly, $\delta^{13}C$ values are higher in cool phase years for the Gulf of Alaska (Figure \@ref(fig:coefres)D) and lower in cool phase years for Washington (Figure \@ref(fig:coefres)B). This suggests there is lower phytoplankton growth in Washington and higher in Gulf of Alaska in cool phase years. This is contrary to results of previous studies, assuming 1) higher $\delta^{13}C$ values represent higher growth rates and 2) PDO is inhibiting growth at the base of the food web and indirectly constraining higher trophic levels such as salmon [@Mantua1997]. It is likely the relationship between PDO, salmon production, and $\delta^{13}C$ values of harbors seals is instead caused by phytoplankton community structure constraining higher trophic levels rather than growth. 

Common temporal trends in harbor seal stable isotopes and ocean condition empirically derived from the GPDFA analysis (Figures 6 & 7) show changes in biogeochemical cycling and food web-assimilated production in recent decades that are associated with climatic variables. Since 1975, shared trends in environmental time series and stable isotope data in the Gulf of Alaska are above average for temperature, discharge, and NPGO and below average for assimilated $\delta^{13}C$ values (as indicated by its negative loadings; Figure \@ref(fig:GDFAAK)). Trends 2 and 3 in the Gulf of Alaska (Figure \@ref(fig:GDFAAK)) show a distinct change in environmental indices starting in 1988. Loadings on these trends were higher for environmental indices than stable isotope data, suggesting a decoupling of environment-food web relationship in the region starting around 1988, which has also been observed between climate regimes and fish species [@Litzow2020]. This environmental-food web decoupling was not observed in Washington (Figure \@ref(fig:GDFAWA)) in our study or others [@Litzow2020]. 

###Using stable isotopes as food web indicators

Previous research has shown lower trophic levels are sensitive to environmental variation in bottom-up drivers of productivity [@Frank2015; @Jennings2010], but few studies have demonstrated how the impact of these changes span entire food webs on long time scales. By applying CSSIA to museum specimens of a generalist predators, we provide a novel piece of the ecological puzzle not previously available. First, these data provide a measure of changing nitrogen resources and phytoplankton dynamics that are spatially and temporally integrated for food web resource assimilation, rather than measuring the availability of inorganic nutrients or lower trophic level biomass and assuming an associated food web response. Dominant species of marine zooplankton exhibit selective foraging, particularly when resources are highly available [@Bi2020; @Jungbluth2017; @Boersma2015] thus discrete measures of resources are not necessarily representative of what is utilized by the food web. Second, studies directly measuring primary production are often temporally limited to short time scales and recent decades. CSSIA of historic specimens allows for retrospective analyses that span long time scales [@McMahon2019; @Sherwood2011] and thus identify long-term environmental forcing on food webs. 

Despite these benefits, CSSIA (and stable isotope analysis data more generally) is limited in its ability to discern different mechanistic processes for isotopic enrichment in observational studies. Multiple mechanisms of fractionation often operate in tandem (Figure 2.1) and can be both additive and subtractive. For example, both the isotopic composition of dissolved inorganic nitrogen sources (primarily $NO_3^-$, but also urea and $NH_4^+$) and the relative uptake of these sources impact the isotopic composition of nitrogen in primary producers [@Graham2010; @Ohkouchi2017]. As a result, these data on their own are limited in their ability to track exact mechanisms of fractionation and specific biogeochemical changes through time or space. Regardless, stable isotope signatures of nitrogen from source amino acids and bulk carbon can be used to trace variations in nitrogen sources at the base of the food web [@Sherwood2014; @delaVega2019] and changes in phytoplankton dynamics (i.e., production) [@delaVega2019; @Lorrain2020] broadly. In addition, CSSIA of carbon is also emerging as reliable proxy for phytoplankton community composition [@Larsen2013]. We also assume a constant and small trophic enrichment factor for both bulk $\delta^{13}C$ and  $\delta^{15}N_{Phe}$ values. While trophic enrichment in $\delta^{13}C$ and $\delta^{15}N_{Phe}$ values is minimal [@Hobson1996; @Bocherens2003; @Germain2013; @Ohkouchi2017], and thus unlikely to impact overall correlations between datasets, it can produce enriched absolute isotope values and increased variation between observations, which was not accounted for in this study. Nonetheless, ours is among a number of supporting studies that show food webs are impacted by changing environmental conditions in the northeast Pacific [@Cunningham2018; @Puerta2019; @Stachura2014]. 
	
Climate change will alter nutrient distributions and primary production throughout the worlds’ oceans [@Marinov2010; @Kwiatkowski2017]. Based on analysis of historical patterns of consumer isotopic variation with environmental forcing, we anticipate there will be region-specific spatial variability in how primary production and its dependent food webs respond to environmental change throughout the northeast Pacific over the next century. As environmental conditions (i.e., sea surface temperature, discharge, anthropogenic nitrogen) continue to change, so will resources available to and assimilated by food webs. Given both resource availability and community composition of resources impact the function and stability of food webs [@Narwani2012] it is likely that ecosystem interactions will change in response to environmentally induced shifts in resources. Understanding dynamics influencing food web responses to their environment is important, as it provides information useful for predicting climate change impacts to aquatic resources and the communities and economies that depend on them.

\clearpage

##Tables

**Table** \@ref(tab:ranges): Range of $\delta^{13}C$ and $\delta^{15}N_{Phe}$ values observed in harbor seals for each of the five northeast Pacific subregions. 

```{r ranges,echo=FALSE}
supp1 <- read.csv("data/Ch2/Table3.csv")
 supp1<-  kableExtra:::kable_styling(knitr::kable(supp1, "latex",col.names = gsub("[.]", " ", names(supp1)), 
                  caption = "Ranges of stable isotope data",
                  longtable = TRUE, booktabs = TRUE
                  ),font_size = 8)
kableExtra::column_spec(supp1, column=2:4, width = "10em")
```

\clearpage

\begin{landscape}
Table 2.2: Environmental datasets. SST data was obtained from NOAA ERSST V5 data provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their Web site at https://www.esrl.noaa.gov/psd/ 


```{r envvar,echo=FALSE}
supp1 <- read.csv("data/Ch2/Table1.csv")
 supp1<-  kableExtra:::kable_styling(knitr::kable(supp1, "latex",col.names = gsub("[.]", " ", names(supp1)), 
                  caption = "Environmental Datasets",
                  longtable = TRUE, booktabs = TRUE
                  ),font_size = 8)
kableExtra::column_spec(supp1, column=2:4, width = "20em")
```
\end{landscape}


**Table** \@ref(tab:reduced): Reduced time series  for linear models by regions.


```{r reduced,echo=FALSE}
supp1 <- read.csv("data/Ch2/Table2.csv")
 supp1<-  kableExtra:::kable_styling(knitr::kable(supp1, "latex",col.names = gsub("[.]", " ", names(supp1)), 
                  caption = "Reduced Time Series",
                  longtable = TRUE, booktabs = TRUE
                  ),font_size = 8)
kableExtra::column_spec(supp1, column=2:4, width = "15em")
```

\clearpage


**Table** \@ref(tab:ttestN): Pairwise t-test by sub region with bonferroni correction and pooled standard deviation for $\delta^{15}N_{Phe}$.


```{r ttestN,echo=FALSE}
supp1 <- read.csv("data/Ch2/Table4.csv")
 supp1<-  kableExtra:::kable_styling(knitr::kable(supp1, "latex",col.names = gsub("[.]", " ", names(supp1)), 
                  caption = "Nitrogen Phenylalanine T-Test",
                  longtable = TRUE, booktabs = TRUE
                  ),font_size = 8)
kableExtra::column_spec(supp1, column=2:4, width = "10em")
```

\clearpage

**Table** \@ref(tab:ttestC): Pairwise t-test by sub region with bonferroni correction and pooled standard deviation for $\delta^{13}C$.


```{r ttestC,echo=FALSE}
supp1 <- read.csv("data/Ch2/Table5.csv")
 supp1<-  kableExtra:::kable_styling(knitr::kable(supp1, "latex",col.names = gsub("[.]", " ", names(supp1)), 
                  caption = "Bulk Carbon T-Test",
                  longtable = TRUE, booktabs = TRUE
                  ),font_size = 8)
kableExtra::column_spec(supp1, column=2:4, width = "10em")
```

\clearpage

##Figures

\begin{landscape}

Figure 2.1: Mechanisms of environmentally induced changes in resources (A-D) assimilated into stable isotope ratios of primary producers (1-2), which are conserved when assimilated into higher trophic levels in the food web (3).\newline 

\begin{figure}[h]
\centering
  \includegraphics[width=1.1\textwidth]{figure/Ch2/Figure1.pdf}
  \caption{Mechanisms of stable isotope change}
  (\#fig:theo)
\end{figure}
\end{landscape}
\clearpage


**Figure** \@ref(fig:map): Spatial and temporal distributions of northeast Pacific harbor seal specimens by subregion analyzed for $\delta^{15}N_{Phe}$ and bulk $\delta^{13}C$ values. Subplot colors correspond to map locations and x-axis (years) is the same for each subplot.\newline 

\begin{figure}[h]
\centering
  \includegraphics[width=1.2\textwidth]{figure/Ch2/Fig2_Map.pdf}
  \caption{Distribution of harbor seal specimens}
  (\#fig:map)
\end{figure}
\clearpage

**Figure** \@ref(fig:dist): Variability in $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values based on sub region and sex. * denotes a significant difference in isotopic signature between males and females for that region (colors correspond to \@ref(fig:map)). 
\newline 

\begin{figure}[h]
\centering
  \includegraphics[width=1\textwidth]{figure/Ch2/Figure3_SexLocation.pdf}
  \caption{Distribution of harbor seal specimens}
  (\#fig:dist)
\end{figure}

\clearpage

**Figure** \@ref(fig:hiermod): Relationship between nitrogen sources ($\delta^{15}N_{Phe}$) and primary production ($\delta^{13}C$) assimilated into the food web for A. a single linear model for the combined data across the northeast Pacific and eastern Bering Sea and B. a mixed effects model with random slope and intercept by sub region (colors correspond to  \@ref(fig:map)). 
\newline 

\begin{figure}[h]
\centering
  \includegraphics[width=1\textwidth]{figure/Ch2/hiermod.pdf}
  \caption{Hierarchical $\delta^{15}N_{Phe}$ and $\delta^{13}C$ Models}
  (\#fig:hiermod)
\end{figure}

\clearpage


**Figure** \@ref(fig:coefres): Coefficients of environmental covariates for models with relative support ($\Delta AIC_c$ < 2) for harbor seal $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values in three regions of the northeast Pacific: Washington, Gulf of Alaska, and the eastern Bering Sea. Color indicates model support based on AIC~c~ weight, points are the coefficient estimates for each environmental covariate included in an individual model, and bars show two standard deviations from the coefficient estimate. 
\newline 

\begin{figure}[h]
\centering
  \includegraphics[width=0.9\textwidth]{figure/Ch2/Figure5_CoefPlot.pdf}
  \caption{Coefficients of Evironmental Covariates}
  (\#fig:coefres)
\end{figure}

\clearpage

**Figure** \@ref(fig:GDFAAK): Common trends in environmental condition and food web assimilated stable isotope values for the regional Gulf of Alaska gaussian-dynamic factor analysis model. The solid lines represent the modelled trends, where 0 is the long-term average and 1 and -1 represent the maximum and minimum possible values respectively; the dash line is the 90% credible interval. Factor loadings can be interpreted as coefficients, representing the strength of association between the modelled trend and each observed environmental time series (colors represent a priori driver category). Values close to 0 mean the observed time series did not correlate to the corresponding trend, while values close to 1 show the observed time series closely matched the modelled trend.  Negative loadings indicate an inverse relationship between the observed time series and modelled trend. Stable isotope times series are modelled separately for the northern (N. $\delta^{15}N_{Phe}$; N. $\delta^{13}C$) and southeast (S. $\delta^{15}N_{Phe}$; S. $\delta^{13}C$) subregions.  
\newline 

\begin{figure}[h]
\centering
  \includegraphics[width=0.9\textwidth]{figure/Ch2/Figure6_AK.GDFA.pdf}
  \caption{Alaska GDFA Results}
  (\#fig:GDFAAK)
\end{figure}

\clearpage

**Figure** \@ref(fig:GDFAWA): Common trends in environmental condition and food web assimilated stable isotope values for the regional Washington gaussian-dynamic factor analysis model. Stable isotope times series are modelled separately for the coastal (C. $\delta^{15}N_{Phe}$; C. $\delta^{13}C$) and Salish Sea (S.S. $\delta^{15}N_{Phe}$; S.S. $\delta^{13}C$) subregions. See \@ref(fig:GDFAAK) caption for further interpretation.  
\newline 

\begin{figure}[h]
\centering
  \includegraphics[width=0.9\textwidth]{figure/Ch2/Figure7_WA.GDFA.pdf}
  \caption{Washington GDFA Results}
  (\#fig:GDFAWA)
\end{figure}

\clearpage

**Figure** \@ref(fig:Month): Analysis of a) $\delta^{15}N_{Phe}$ and b) $\delta^{13}C$ values by month. For both models, s(month) p > 0.1 indicating no seasonality of harbor seal bone collagen stable isotope signature.  
\newline 

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\textwidth]{figure/Ch2/figures2.png}
  \caption{Bone collagen stable isotope seasonality analysis}
  (\#fig:Month)
\end{figure}

\clearpage

**Figure** \@ref(fig:Length): Analysis of a) $\delta^{15}N_{Phe}$ and b) $\delta^{13}C$ values by length. For both models, there was no significant slope (p>0.1)
\newline 

\begin{figure}[h]
\centering
  \includegraphics[width=0.8\textwidth]{figure/Ch2/FigureS2.pdf}
  \caption{Bone collagen stable isotope length analysis}
  (\#fig:Length)
\end{figure}

\clearpage

**Figure** \@ref(fig:TSphe): Bone collagen $\delta^{15}N$ values of phenylalanine from archival harbor seal specimens collected in the northeastern Pacific in five subregions.
\newline 

\begin{figure}[h]
\centering
  \includegraphics[height=0.8\textwidth]{figure/Ch2/FigureS3.pdf}
  \caption{Time series of $\delta^{15}N_{Phe}$ data}
  (\#fig:TSphe)
\end{figure}

\clearpage


**Figure** \@ref(fig:nonSuess): Bone collagen bulk $\delta^{13}C$ values of archival harbor seal specimens collected in the northeastern Pacific in five subregions.These values are not corrected for the Suess effect.
\newline 

\begin{figure}[h]
\centering
  \includegraphics[height=0.8\textwidth]{figure/Ch2/FigureS4.pdf}
  \caption{Time series of $\delta^{13}C$ data}
  (\#fig:nonSuess)
\end{figure}

\clearpage

**Figure** \@ref(fig:Suess): Bone collagen bulk $\delta^{13}C$ values of archival harbor seal specimens collected in the northeastern Pacific in five subregions.These values are corrected for the Seuss effect.
\newline 

\begin{figure}[h]
\centering
  \includegraphics[height=0.8\textwidth]{figure/Ch2/FigureS5.pdf}
  \caption{Time series of $\delta^{13}C$ data corrected for Suess effect}
  (\#fig:Suess)
\end{figure}

\clearpage

**Figure** \@ref(fig:bulkN): Bone collagen bulk $\delta^{15}N$ values of archival harbor seal specimens collected in the
northeastern Pacific in five subregions.
\newline 

\begin{figure}[h]
\centering
  \includegraphics[height=0.8\textwidth]{figure/Ch2/FigureS6.pdf}
  \caption{Time series of bulk $\delta^{15}N$ data}
  (\#fig:bulkN)
\end{figure}

\clearpage

**Figure** \@ref(fig:linresid): Residuals for the model with the most support (\@ref(fig:coefres)) plotted by year. A trend in model residuals would indicate environmental variables do not account for all temporal variation in harbor seal $\delta^{15}N_{Phe}$ and $\delta^{13}C$ values.
\newline 

\begin{figure}[h]
\centering
  \includegraphics[height=0.85\textwidth]{figure/Ch2/FigureS7.pdf}
  \caption{Residuals trends for linear models with the most support}
  (\#fig:linresid)
\end{figure}

\clearpage


Figure 2.15: Model residual plots for the models with the most support from the candidate model set. Note: EBS phenylalanine is an intercept only model, GOA phenylalanine only contains a 2-factor location as a covariate.
\newline 

\begin{figure}[h]
\centering
  \includegraphics[height=0.8\textwidth]{figure/Ch2/FigureS8.pdf}
  \caption{Residuals for linear models with the most support}
  (\#fig:linresid2)
\end{figure}
