# Recent divergent changes in Alaskan pinniped trophic position detected using compound-specific stable isotope analysis

##Abstract

Over the past century Alaskan pinnipeds have experienced dramatic changes in abundance, but these changes have been highly variable across species and region. In recent decades, changes in atmospheric forcing and sea surface temperature have been particularly pronounced in the Gulf of Alaska and eastern Bering Sea, impacting the food webs in which Alaskan pinnipeds forage. We used compound-specific stable isotope analysis of nitrogen in amino acids to estimate historic and modern trophic position of harbor seals and Steller sea lions in the Gulf of Alaska and Bristol Bay. We applied a Bayesian hierarchical framework to determine whether shared trends through time exist across pinnipeds (classified by species and region) on decadal scales. Model results identified both shared trends through time and classification-specific decadal changes in pinniped trophic position. The largest change in trophic position occurred in the 2000s and 2010s and was observed in both Steller sea lions and harbor seals in the Gulf of Alaska, but not harbor seals in Bristol Bay or Iliamna Lake. Divergent trophic position patterns in the 2000s were identified in the western stock of Steller sea lions, which increased in trophic position, and sympatric harbor seals in the northern Gulf of Alaska, which decreased in trophic position. Our results indicate that these species have begun exploiting distinct trophic niches or experiencing unique food web conditions in recent decades in the Gulf of Alaska, likely in response to recent climate-induced ecological change in the region.

##Introduction
Over the past century, pinniped populations in the northeast Pacific Ocean have experienced changes in adult and pup abundances [@Muto2020]. Understanding specific drivers of these population trends is important for management, as multiple stocks have been listed as threatened or endangered over the past two decades [@Muto2020]. The observed population dynamics have also corresponded with shifts in both the physical and ecological marine environment, which frequently occur simultaneously. As a result, disentangling drivers of population trends is complex, as multiple factors (environmental conditions, prey availability, anthropogenic disturbances) can change in tandem and potentially act synergistically on pinniped populations. 

Data on long-term trends in trophic position across regions, species, and populations is one potential way to assess how food web changes have impacted pinnipeds in Alaska. This approach can identify how broad shifts in foraging ecology correspond to changes in abundance and population dynamics. More specifically, examining trophic position during periods of declining versus increasing predator abundance can provide insight into whether foraging behavior and prey availability are important drivers of population dynamics. In this study, we aim to identify whether common temporal trends in trophic ecology exist across harbor seals (*Phoca vitulina*) and Steller sea lions (*Eumetopias jubatus*) and their locations by deriving 70-years of trophic position data from compound-specific stable isotope analysis (CSIA) of museum specimens.

Following climatic changes in the 1970s that altered ocean currents and sea surface temperature [@Hare2000], most Gulf of Alaska and Bering Sea pinniped populations experienced declines that persisted through the 1990s [@Muto2020]. However, these responses differed across populations and species. For example, the western stock of Steller sea lions (located west of 144°W) decreased from approximately 240,000 animals in the late 1970s to 50,000 in 2000 [@Burkanov2005]. Similarly, harbor seal populations in Prince William Sound and Glacier Bay declined by approximately 60% between the 1980s and 2000 [@Frost1999; @Womble2010]. In contrast, the eastern stock of Steller sea lions (located east of 144°W) increased by 3-4% per year over the same time period (Figure 2) [@Muto2020; @Pitcher2007]. More recently, atmospheric circulation anomalies in the northeast Pacific Ocean have resulted in unprecedently warm sea surface temperatures during the past decade [@Walsh2018] and this environmental shift has altered fish abundances [@Bond2015; @Litzow2020]. For example, the unprecedented marine heatwave that occurred in 2014 - 2016 triggered dramatic ecosystem change, including a 71% decline in Pacific cod in the Gulf of Alaska [@Barbeaux2020]. Declines in phytoplankton biomass, forage fish abundance, and changes in community structure as a whole were also observed [@Suryan2021]. During this recent period of environmental change, many pinniped populations have experienced increases or stabilization of population abundance [@Muto2020] (Figure 2), although declines in some Gulf of Alaska Steller sea lion populations were observed following the marine heat wave [@Suryan2021].

These variable changes in Alaskan pinniped populations over the past 50 years cannot be attributed to a single cause, as multiple environmental, anthropogenic, and ecological factors have changed simultaneously. For example, the rapid decline of the western stock of Steller sea lions between the 1970s and 1990s has been attributed to myriad factors, including change to the physical environment, competition with fisheries for common prey, predation, disease, and human-caused mortality [@Atkinson2008]. Glacier Bay harbor seal populations have primarily, but not exclusively, been impacted by the decline of sea ice, which provides a majority of their haulout sites [@Womble2010]. Population declines have also been associated with increased numbers of tour vessels, particularly in glacier fjords that provide important nursing and whelping habitat [@Jansen2015; @Matthews2016]. The differences in pinniped population trends across the Gulf of Alaska and Bering Sea suggest varied environmental and ecological drivers underlying these dynamics. Interestingly, harbor seals and Steller sea lions that occur in the same geographic region (sympatric) have experienced different population trends over similar time period (Figure 2). Identifying trophic position trends through time that are shared, compared to changes that only impact a specific species or region, can elucidate how widescale ecological forcing versus localized change influence top predators and potentially explain variable population abundance trends.

Both harbor seals and Steller sea lions exhibit generalist, piscivorous foraging strategies, although differences in foraging range, body size, and diet exist. Adult harbor seals have high site fidelity, opportunistically forage 5 - 10 km from haulout sites and at depths < 200 m [@Lance2012; @Lowry2001], and weigh up to 300 pounds. Steller sea lions are central place foragers known to migrate to prey aggregations on the continental shelf and oceanographic boundary zones [@Sinclair2002; @Womble2006]. Foraging trips can last 1-3 days [@Maniscalco2006] with average distances of 133 km for adult females [@Merrick1997], although foraging trips are shorter in the breeding season [@Maniscalco2006]. Adult females can weigh up to 800 pounds whereas adult males can exceed 2,500 pounds, indicating a higher energetic demand compared to harbor seals. Diet studies of Steller sea lions and harbor seals are spatially and temporally limited, and primarily utilize scat samples. In the Gulf of Alaska, gadids, cephalopods, and forage fishes are prevalent in both harbor seal and Steller sea lion diet [@Sinclair2002; @Geiger2013], whereas salmonids are also important for harbors seals in Bristol Bay and Iliamna lake [@Hauser2008]. 

Stable isotopes have been used to reconstruct historical differences in diet and trophic position in Alaskan pinnipeds [@Hobson1997; @Hirons2001; @Brennan2019]. These previous studies utilized bulk stable isotope analysis exclusively and were therefore limited in their inferential strength. Differences in the bulk ^15^N/^14^N of consumer tissues can indicate either a trophic level change of the consumer or a change in nitrogen resources at the base of the food web. The specific cause of the isotopic variation cannot be ascertained from consumer bulk stable isotope values unless the data are paired with temporal information on ^15^N/^14^N in primary producers. Lack of consistent, concurrent sampling of nitrogen stable isotope composition of primary producers therefore presents a challenge for previous long-term studies of the trophic dynamics of consumers from bulk stable isotope data. CSIA data address this challenge, as amino acids exhibit two distinct patterns in isotopic enrichment: trophic amino acids (i.e., glutamic acid, alanine, proline) become enriched in ^15^N with each trophic transfer and source amino acids (i.e., phenylalanine) show minimal change and thus are reflective of the base of the food web [@McClelland2002; @Chikaraishi2009; @Ohkouchi2017]. With the ability to internally correct for expected changes in ^15^N/^14^N at the base of the food web [@Feddern2021; @McMahon2021], CSIA allows for a more robust retrospective analysis of consumer trophic dynamics on decadal and century scales.
	 
The objective of this work is to describe and compare changes in trophic ecology for Alaskan pinnipeds throughout the past century and investigate trophic position differences for sympatric populations. We apply hierarchical Bayesian analyses to 70 years of trophic position data derived from CSIA from pinnipeds (harbor seal and Steller sea lion) in the Gulf of Alaska, Bristol Bay, and a small population of freshwater harbor seals in Iliamna Lake, Alaska which is adjacent to Bristol Bay. We build on previous research examining pinniped nitrogen stable isotope composition [@Hobson1997; @Hirons2001; @Misarti2009; @Brennan2019] by adding two decades of data to the record (2000s and 2010s) and incorporating a broad spatial scope (Bristol Bay, Iliamna Lake, Gulf of Alaska). Additionally, by analyzing nitrogen stable isotopes derived from amino acids, we were able to control for known changes in nitrogen resources and phytoplankton composition at the base of the food web that can confound trophic position interpretations from bulk stable isotope data collected over decadal scales [@Feddern2021]. Furthermore, by comparing trophic position dynamics across species and region through time, regional and location-specific ecological responses to a changing ecosystem can be identified. 

##Methods

###Sample collection and analysis

Samples were obtained using methods described in @Feddern2021. Briefly, harbor seal and Steller sea lion bones were sampled from specimens curated at the University of Alaska Museum of the North (Supplementary Information Table S1). Specimens were treated by maceration in warm water and soaked in a dilute ammonia solution then stored in acid free boxes. Adult specimens were sampled exclusively to avoid dietary differences between adults and juveniles. Specimens were classified based on species and region. We prioritized long-term temporal coverage in four regional classifications of harbor seals (Iliamna Lake, southeast Gulf of Alaska, northern Gulf of Alaska, eastern Bering Sea) and two regional classifications of Steller sea lions (eastern and western stocks) for a total of 6 species x region classifications. Specimens were extremely limited for the eastern Steller sea lion stock (n = 2) and Iliamna Lake harbor seas (n = 3). We also prioritized specimens with sex and age identifications, but these data were not available for some specimens. A total of 106 harbor seal and 21 Steller sea lion specimens were sampled representing the 1950s to 2010s (Figure 1).

Steller sea lions were classified according to the National Oceanic and Atmospheric Administration’s (NOAA) distinct population segments, where Steller sea lions east of 144°W are considered the eastern stock and west of 144°W are considered the western stock (Figure 1). NOAA has identified twelve stocks of harbor seals in Alaska and, due to limitations of archived specimens, harbor seals were not able to be classified according to NOAA stocks. Instead, they were classified based on their range relative to the Steller sea lion stocks and utilization of marine versus freshwater habitats. Harbor seals that were west of 144°W, which included samples from the Prince William Sound and Cook Inlet/Shelikof Strait stocks (Figure 1), were classified as northern Gulf of Alaska harbor seals. Harbor seals that were located east of 144°W, which included samples from the Glacier Bay/Icy Strait, Sitka/Chatham Strait, Lynn Canal/Stephens Passage, Dixon/Capes Decisions, and Clarence Strait stocks (Figure 1), were classified as southeast Gulf of Alaska harbor seals. The Bristol Bay harbor seal stock was divided into two classifications, Bristol Bay referring to marine harbor seals, and Iliamna Lake referring to freshwater harbor seals (Figure 1). This allowed for comparison of three pairs of geographically overlapping classifications: western stock of Steller sea lions and northern Gulf of Alaska harbors seals, eastern stock of Steller sea lions and southeast Gulf of Alaska harbor seals, and Bristol Bay and Iliamna Lake harbor seals.  

###Trophic position calculation

Bone collagen within the samples was decalcified, acid hydrolyzed, derivatized and analyzed for compound-specific stable isotope analysis (CSIA) of nitrogen ($\delta^{15}N$) for 12 individual amino acids following the protocol described in @Feddern2021. $\delta^{15}N$ was measured as:

\begin{equation} 
  \delta^{15}N ( \textperthousand vs. air) =   
  [(\frac{^{15}N/^{14}N_{Sample}}{^{15}N/^{14}N_{Air}} -1)*1000]
  (\#eq:deltN2)
\end{equation} 

Collagen samples were measured in triplicate with a laboratory standard containing a 12 amino acid mixture of known isotopic composition. Full analytical details are described in Appendix S1.

Trophic position was calculated using a harbor seal-specific trophic discrimination factor (difference in ^15^N/^14^N between trophic and source amino acids in consumers for a trophic transfer; @Germain2013). This approach assumed trophic discrimination factors (TDF) derived from controlled feeding studies of harbor seals were similar to Steller sea lions. Applying a "multi-TDF" approach that combines both average and taxa-specific TDF can improve trophic position estimates in marine predators including pinnipeds [@Germain2013; @McMahon2019]. The following equation was used to determine the trophic position of each sampled individual: 

\begin{equation} 
Trophic Position =   
  \frac{\delta^{15}N_i - \delta^{15}N_o - TDF_{(i-o),j} - \overline{\beta}_{(i-o)}}{\overline{TDF}_{(i-o)}}+2
  (\#eq:TP)
\end{equation} 

where, $\delta^{15}N_i$ is the measured stable isotope composition of a trophic amino acid i in a sample and $\delta^{15}N_o$ is the stable isotope composition of a source amino acid o in a sample. $\overline{TDF}_{(i-o)}$  is the mean difference between given trophic amino acid *i* and source amino acid *o* across all consumers described in @Nielsen2015. $TDF_{(i-o), j}$ is the trophic discrimination factor between trophic amino acid *i* and source amino acid *o* from a controlled feeding study of a specific consumer *j*; here we use harbor seals from @Germain2013 (Table 1). $\overline\beta_{(i-o)}$ is the mean difference in $\delta^{15}N$ across aquatic phytoplankton between a specific trophic amino acid *i* and source amino acid *o* [@Nielsen2015; Table 1]. @Nielsen2015 also determined using multiple amino acids to estimate trophic position improves precision. Therefore, we used multiple trophic amino acids *i* (alanine, glutamic acid, aspartic acid and proline) and one source amino acid *o* (phenyalanine) to calculate trophic position (Table 1). These amino acids were chosen based on their prevalence in previous studies to derive parameters for equation 2, and their concentrations in bone collagen (see Appendix S1).

###Model framework
Sex was considered as a predictor for trophic position, however, sex metadata were not available for all specimens. In order to evaluate difference in trophic position by sex, we fit linear statistical models to each individual trophic amino acid, by classification (species x region). These models took the following form:

\begin{equation} 
y_i = \alpha + \boldsymbol\beta x_i + \epsilon_i, \epsilon \sim N(0,\sigma)
  (\#eq:linsex)
\end{equation} 

where, $y_i$ is trophic position for an individual amino acid and $\boldsymbol\beta$ is a vector of coefficients for the predictor, in this case sex, and $\epsilon$ are residual errors assumed to be normally distributed with mean 0 and standard deviation  $\sigma$. There was not sufficient metadata for the eastern stock Steller sea lion population or the Iliamna Lake population and these two classifications were omitted from this analysis.

A Bayesian hierarchical mixed effects model was used to identify decadal change across pinniped classifications (species x region), and the degree to which these changes were shared by testing the effects of classification, decade, and a classification-decade interaction as either population-level (fixed) or group-level (random) effects (see candidate models in Table 2). Hierarchical models share information across ‘groups’ to identify common responses, which refers to both decade and classification in this study. The interaction term allows for increased flexibility, letting each classification have slight departures from the group-level means. The mean and variance of pinniped trophic position for each region-species classification and decade were estimated using a generalized linear Bayesian hierarchical model with decade, population, and trophic amino acid as predictors:

\begin{equation} 
y_i = \boldsymbol\alpha + \boldsymbol\beta x_i + \epsilon_i, \epsilon \sim N(0,\sigma_y)
  (\#eq:linsex)
\end{equation}

\begin{equation} 
\boldsymbol\alpha_{k=1:k} \sim N(\mu_{\alpha,k},\sigma_{\alpha,k})
  (\#eq:linsex)
\end{equation}

where, for data point *i*, $\boldsymbol\beta$ is a vector of coefficients for the unpooled predictors (fixed effects, Table 2) and $\boldsymbol\alpha$ is a vector of coefficients for the partially pooled group-level predictors (random effects, Table 2) for group *k* (amino acid, decade or classification). At minimum, the $\boldsymbol\alpha$ included a random term for the amino acid corresponding to data point *i*, and depending on the model included up to a total of 4 random effects (also effects of decade, classification, and their interaction, Model 6 in Table 2). For each random effect included, $\mu_{α,k}$ and  $\sigma_{α,k}$ are hyperparameters representing the mean and standard deviation of group-level effects on trophic position, for random effect *k*. For models with more than one random effect, we assumed the deviations to be independent and uncorrelated. We considered models that included decade, classification, and the interaction between decade and classification either as fixed or random effects (e.g. Model 4 v Model 6, Table 2), but did not consider models that included both as fixed and random (Table 2). Parameter estimates were obtained using the brms package (@Burkner2017, version 2.14.4) in R (R Core Development Team 2021, version 3.6.2), which implements a Hamiltonian Monte Carlo sampler and its extension no-U-turn sampler [@Hoffman2014] through Stan (Stan Development Team 2020). Minimally informative priors were used for random effects (normal distributions with a mean of 0 and variance of 10) and fixed effects (Student’s t-distribution with a mean of 0, standard deviation of 2.5 and 3 degrees of freedom). Trophic amino acid was included as a random effect for all models (Table 2). Selection of the best models (Table 2) given the data was based on approximate leave-one-out cross-validation (LOOIC) using the loo package (@Veharti2017, version 2.4.1).

##Results
We found no differences between the average male and female pinniped trophic position over the 50-year study period (Figure 3) for the four tested species-region classifications. This finding was consistent for all trophic amino acids-source amino acid pairs (Figure 3). Based on glutamic acid trophic position estimates, both western stock Steller sea lions (2.6 ± 0.5; mean ± sd) and eastern stock Steller sea lions (2.7 ± 0.16) had similar trophic positions. Harbor seals in the Gulf of Alaska foraged higher in the food web than their Steller sea lion counterparts (Figure 3). Harbor seals in the southeast region had a higher trophic position on average than any other pinniped in this study (3.5 ± 0.3) but were similar to harbor seals in the northern region (3.3 ± 0.5). Bristol Bay (3.1 ± 0.4) and Iliamna lake (3.0 ± 0.3) harbor seals had a lower trophic position than their Gulf of Alaska counterparts on average (Figure 3).

###Common trends in Alaskan pinniped trophic position

The best performing model (Table 2, model 6) of pinniped trophic position included both species-region classification and decade as random effects (shared trends) along with an interaction between population and decade (Table 2). Based on the support for decade and classification to be included as group-level effects, these data support consistent differences between classifications over time, as well as differences between trophic position for all classifications. The supported interaction between population and decade (Table 2) indicates distinct decadal changes in trophic position for species-region classifications exist. The model that included decade, classification, and the interaction between decade and classification as fixed effects (model 4) was also supported based on the models LOOIC (Table 2). Therefore, the inclusion of the interaction term was more important for improving model performance than inclusion of decade and classification as fixed versus random effects.

There were consistent differences in trophic position that varied by species and ocean basin for the model with the most support. Harbor seals in the Gulf of Alaska had higher trophic position than their Steller sea lion counterparts. The mean difference of the posterior distributions indicated southeast Gulf of Alaska harbor seals have historically fed at 0.32 [-0.01, 0.61] (highest density 80% credible interval) trophic levels higher than sympatric eastern stock Steller sea lions (Figure 4). Similarly, the mean difference of posterior distributions showed northern Gulf of Alaska harbor seals fed 0.28 [-0.03, 0.50] trophic levels higher than the sympatric western stock Steller sea lions. Within the Gulf of Alaska, the posterior distributions for trophic position overlapped 39% between harbor seals and Steller sea lions in both the eastern and western regions (Figure 4). Iliamna Lake harbor seals have historically fed at a lower trophic level (mean posterior difference 0.16 [-0.11, 0.41]) than harbor seals in Bristol Bay, but these two classifications have 66% overlap of the group-level posterior distributions for trophic position (Figure 4). The 80% credible intervals included 0 for most region-species classifications thus the posterior probabilities support marginal evidence for consistent differences in trophic position between classifications. Regardless, the differences in posterior means were large, although the distributions were wide. 

There were no consistent decadal differences in trophic position across the region-species classification (Figure 5). Pinniped trophic position in the 2000s was slightly higher for all classifications (mean posterior difference 0.03 [-0.09, 0.16]) on average compared to 1990 and the posterior distributions for 1990 and 2000 had an 85% overlap (Figure 5). Similarly, posterior distributions in between 2000 and 2010 had a mean difference of -0.1 [-0.27, 0.08] with a 65% overlap (Figure 4). Overall, decadal differences in pinniped trophic position through time were smaller than the region-species classification effects and were likely ecologically inconsequential.

###Spatial and temporal differences in pinniped trophic structure

Distinct decadal changes in trophic position were observed for each species-region classification and varied more than the shared decadal changes (Figure 6) as indicated by the decade-classification interaction. Most, but not all, pinniped classifications experienced substantial trophic level change in 2000 or 2010 but the magnitude and direction of this change varied by region-species classification based on the combined effects of decade, classification, and the decade-classification interaction (Figure 6). The recent decadal change in trophic position was most prominent for the western stock of Steller sea lions which had a mean trophic level decrease of 0.43 [-0.25, -0.60] from 1990 to 2000 (a percent decrease of 0.15) with only a 21% overlap between the posterior distributions (Figure 6E). This decline in trophic position remained in the 2010s. A similar decline was observed in the southeast Gulf of Alaska harbor seals. This population experienced relatively stable trophic position from 1960-1990, which then declined on average by 0.31 [-0.19, -0.45] trophic levels in 2000 (33% posterior overlap) (Figure 6C). In contrast, harbor seals in the northern Gulf of Alaska had variable trophic position across decades and had the highest trophic position in 2000 in contrast to their southeast Gulf of Alaska harbor seals and Steller sea lion counterparts (Figure 6B). Data were only available for 2000 and 2010 for the eastern stock Steller sea lions, and trophic position was similar for this population during both of these decades (Figure 6F). Both Bristol Bay and Iliamna Lake harbor seals had relatively stable trophic position from 1950s until 2010s (Figure 6A & B). Bristol Bay harbor seals experienced their lowest trophic level in the 1990s with a 0.24 [-0.54, 0.00] trophic level decrease compared to the 1970s and 2000s, but the posterior distribution still overlapped 54% with other decades (Figure 6A). 

##Discussion

Over the past 70 years, Alaskan pinnipeds have exhibited both common and distinct differences in trophic position across region-species classification on decadal scales (Table 2). While potential drivers of change in trophic position were not tested in this study due to data limitations, our results support a combination of local-scale (i.e., vessel traffic, reduction of glacial ice, local foraging) and regional-scale (i.e., environmental condition, basin-wide prey abundance) changes may be influencing pinniped trophic ecology. Furthermore, the largest decadal changes in pinniped trophic position were distinct for each region-species classification and were most apparent during the most recent two decades (2000s and 2010s). These patterns are more pronounced in the Gulf of Alaska compared to Bristol Bay (Figures 5 & 6). 

###Regional and species trends in harbor seal trophic position 

Both Steller sea lions and harbor seals exhibit generalist foraging patterns [@Lance2012; @Geiger2013]. Diets of Alaskan pinnipeds consist of similar prey species but vary between species, population, and local availability of prey [@Iverson1997; @Hirons2001]. Bulk stable isotope studies in the Gulf of Alaska have shown that Steller sea lions feed lower in the food web compared to harbor seals [@Iverson1997]. Our CSIA analysis confirms the interpretation of these previous studies that isotopic differences can be attributed to trophic position changes and not isotopic shift of basal phytoplankton resources. Both western and eastern stock Steller sea lions have lower trophic position compared to sympatric harbor seal populations but have similar trophic position compared to other populations such as Iliamna Lake. However, despite known differences in both diet [@Sinclair2002; @Geiger2013] and nearshore versus offshore foraging [@Merrick1997; @Lowry2001] between the two species, our results also show historical overlap in trophic position, indicating potential trophic redundancy between harbor seals and Steller sea lions in the Gulf of Alaska.

Harbor seals in Bristol Bay and Iliamna Lake are managed as a single population [@Muto2020] despite lack of evidence of migration by the freshwater population and utilization of different resources [@Brennan2019]. A previous study of strontium and carbon stable isotopes showed Iliamna Lake harbor seals utilize freshwater-derived resources (resident lake fishes), particularly early in life, and exhibit an ontogenetic shift to more marine resources (returning sockeye salmon) later in life [@Brennan2019]. Based on CSIA nitrogen data, Iliamna Lake harbor seals also forage lower in the food web compared to Bristol Bay harbor seals. In addition, both classifications exhibited trophic stability, with the Bristol Bay harbor seals only experiencing a trophic shift in the 1990s relative to the 1960s and 1970s. This coincided with the lowest sockeye salmon returns to Iliamna Lake on record [@Hilborn2003]. Interestingly, the decrease in trophic position in the 1990s occurred simultaneously with decreases in basin wide Bristol Bay harbor seal abundance in the late 1990s, which then stabilized and increased in the 2000s and 2010s (Figure 2). Data were not available for the freshwater harbor seals between 1990 and 2000 and thus it is unclear whether the freshwater population also experienced a trophic position change during the 1990s when sockeye salmon returns were low. While quantitative comparisons to salmon abundance were not made in this study, salmon population abundance and harbor seal trophic ecology and population trends are seemingly interrelated. 

###Recent trophic position changes in the Gulf of Alaska

Trophic position changes were observed in all pinniped classifications in the Gulf of Alaska during the past two decades, although the direction of these changes varied on more local scales. During the past two decades (2000-2020), harbor seals in the Gulf of Alaska have experienced stabilization of most monitored populations following long-term declines that persisted from the 1950s through the 1990s (Figure 2, @Muto2020). During this same time period, harbor seals in both southeast and northern Gulf of Alaska experienced a shift in trophic position that was particularly prominent in the 2000s compared to historic estimates of trophic position (Figure 6B & C). It is possible that the observed trophic position shift may have contributed to the population stabilization of Gulf of Alaska harbor seals, either by an increase in prey availability or opportunistically foraging on a novel prey source. @Gagne2018 observed similar trophic position declines in seabird populations, which were attributed to a shift in diet from fish to squid. A similar dietary shift could explain the observed trophic position shift in southeast Gulf of Alaska harbor seals and western stock Steller sea lions. 

Recent regional change in the Gulf of Alaskan food webs has been well documented in other species and primarily attributed to bottom-up effects of climate [@Barbeaux2020; @Litzow2020]. These region-wide trends likely altered prey availability for pinniped populations in the Gulf of Alaska. How pinniped populations have adapted their foraging ecology, however, indicates regional and species trophic divergence, which could be attributed to either local-scale foraging adaptations or differences in prey availability. Pinniped groups that overlap in space (Figure 1) revealed divergent trends in trophic position between Steller sea lions and harbor seals in recent decades (Figure 6B & E). For example, trophic position of northern Gulf of Alaska harbor seals increased in the 2000s while the western stock of Steller sea lions decreased. For western stock Steller sea lions, this shift also persisted into the 2010s (Figure 6E). Posterior distributions of western stock Steller sea lions and northern Gulf of Alaska harbor seals overlapped by 63% in the 1950s but only overlapped by 3% in the 2000s (Figure 6B & E). The recent change in pinniped trophic position within the Gulf of Alaska coincided with population abundance stabilization, albeit at lower than historical abundance for most populations. This trophic divergence indicates there could be increased competition for resources between northern Gulf of Alaska harbor seals and western stock Steller sea lions resulting in diet adaptations. Similar comparisons were challenging to make for the eastern stock of Steller sea lions and southeast Gulf of Alaska harbor seals due to limitations in historical data for the former. However, trophic position in the 2000s showed a 38% overlap (Figure 6C & E) between the two species, indicating any trophic divergence between them may be less pronounced in this region, if existent. 

The observed divergent trends indicate differences in how Alaskan pinnipeds are adapting to environmental and ecological changes. Trophic position changes from stable isotope data can be accounted for by: 1) prey switching between different species, 2) consuming different sizes of the same prey, or 3) consuming different quality prey. These changes can occur at the consumer level (pinnipeds) or lower in the food web and still be reflected in consumer stable isotope signature and thus trophic position. In recent decades, Pacific salmon and halibut in Alaska have both declined in size [@Holsman2019; @Oke2020]. These changes in size distributions of prey have been attributed to changes in marine mammal populations (Groskreutz et al. 2019) and likely contributed to the observed trophic position declines in western Steller sea lion and southeast Gulf of Alaska harbor seals. In contrast, consuming low-quality prey with lower protein content and greater amino acid imbalance between consumer and prey increases the amino acid trophic enrichment factor of nitrogen [@McMahon2015]. If not accounted for in trophic position equations, this increase in trophic enrichment factor can result in erroneously high trophic position estimates. This may explain the observed increase in estimated trophic position in northern Gulf of Alaska harbor seals where this population may be consuming a greater proportion of lower quality prey (i.e., crustaceans, shrimp, cephalopods) in recent decades rather than feeding on prey species that are higher in the food web. 

###Considerations and limitations for CSIA analyses

The data in this study were limited in sample size primarily due to the availability of archived specimens. As a result, we were not able to discern between known fine scale differences in populations or annual trends. For example, harbor seals in the southeast Gulf of Alaska consist of 13 individual stocks. Due to limitations in the number of archived specimens, these stocks were pooled and analyzed as a single classification despite known differences in genetic structure [@Muto2020]. Given the observed broad range in trophic position of these generalist predators, it is unlikely that inclusion of finer spatial dynamics would have changed the supported model, although variation in temporal trends within a classification may have been identified. Similarly, data were only available for eastern stock Steller sea lions for 2000s and 2010s. As a result, no historical comparisons were possible and the conclusions about this population are tentative. Nonetheless, this dataset offers historic documentation of pinniped trophic position that can be updated with future samples or additional archived specimens.

Trophic position estimates in this study were low compared to known foraging strategies of these pinnipeds. For example, Steller sea lions eat primarily walleye pollock and Atka mackerel [@Hobson1997; @Trites2007], which would indicate a trophic position of 3 or higher. Mean trophic position for Steller sea lions was closer to 2.7 in this study, which is lower than expected based on known foraging ecology. It is common for CSIA to underestimate trophic position of marine predators [@Germain2013; @McMahon2019] and the inclusion of multiple amino acid pairs and a multi-trophic enrichment factor framework did not fully resolve this issue. @Nielsen2015 found trophic position estimates can be highly sensitive to the applied $\beta$ values in equation 2. In our trophic position calculation, we assumed a constant $\beta$ represented by marine diatoms. However, $\beta$ values differ by more than 11 per mille between seagrasses and diatoms [@VanderZanden2013] which has been attributed to differences between vascular and nonvascular plants [@Choi2017, @Ramirez2021]. If vascular plants, such as seagrasses, contribute to the food web in addition to non-vascular algae, the applied β would be too high and would result in underestimation of trophic position of marine consumers [@Choi2017, @Ramirez2021]. Even a 10% contribution of vascular plant-derived nitrogen to the food web would result in an underestimation of 0.2 trophic position. It is likely that vascular plants at least partially contribute to the Alaskan food web, as seagrass beds provide essential habitat and food for many fish species and invertebrates. Consideration for variable $\beta$ values may be helpful in resolving trophic position underestimation of future studies, especially in cases where consumer carbon stable isotope data is available and contributions of seagrasses to the food web are well documented.

###Conclusions and Implications

Marine ecosystems in Alaska are experiencing unprecedented environmental change that has altered abundance and size distributions of many fish species consumed by pinnipeds [@Barbeaux2020; @Holsman2019; @Oke2020; @Suryan2021]. Heterogeneity in diet and foraging locations allow top predators to adjust to availability of resources by altering their foraging. Based on the observed region-species specific changes in trophic position over the past two decades, pinnipeds are experiencing different food web conditions than in the past, even those that occur in similar geographic regions. This may be the result of adapting foraging strategies to exploit other prey resources or a change that is occurring lower in the food web and is measurable in predators. While our results cannot discern between these two mechanisms of trophic level change, we can conclude that recent food web dynamics have impacted pinniped trophic ecology in Alaska. Future responses of pinnipeds to food web change will likely be locally variable between species, even those that occur within similar geographic regions. 

\clearpage


##Tables

**Table** \@ref(tab:paramval): Parameter values for trophic discrimination factors between a trophic amino acid (*i*) and phenylalanine (*o*) for harbor seals ($TDF_{(i-o), j}$), for an average consumer (TDF(i-o) ), and for primary producers ($\beta_{(i-o)}$)  derived from previous studies to apply a multi amino acid framework to equation 2.

```{r paramval,echo=FALSE}
supp1 <- read.csv("data/Ch4/Table1.csv")
colnames(supp1) <- c("Trophic Amino Acid (i)",
                           "$\\overline{\\beta}_{(i-o)}$",
                           "$TDF_{(i-o),j}$",
                           "$\\overline{TDF}_{(i-o)}$")
supp1 <- kableExtra:::kable_styling(knitr::kable(supp1, "latex", escape = FALSE,
caption = "Trophic position parameter values for Equation 2",
                  longtable = TRUE, booktabs = TRUE
                  ),font_size = 8)
 
kableExtra::column_spec(supp1, column=2:4, width = "10em")

```

\clearpage



**Table** \@ref(tab:candmodels): Candidate models for identifying spatial and temporal trophic structure of Alaskan pinnipeds. Assumptions define how the model describes trophic structure with regards to decade and classification and LOOIC describes the support of each candidate models. The best model (6) is italicized.

```{r candmodels,echo=FALSE}
supp1 <- read.csv("data/Ch4/Table2.csv")

 supp1<-  kableExtra:::kable_styling(knitr::kable(supp1, "latex", escape = FALSE, col.names = gsub("[.]", " ", names(supp1)), 
                  caption = "Candidate Models",
                  longtable = TRUE, booktabs = TRUE
                  ),font_size = 8)
supp1 <- kableExtra::row_spec(supp1,6,italic=T)
kableExtra::column_spec(supp1, column=2:4, width = "13em")

```

\clearpage